---
title: "Week 03 - Interactivity and Animation"
format: html
sidebar: false
date-modified: last-modified
---

# 3 Programming Interactive Data Visualisation with R

## 3.1 Introduction

The following content is created by following the tutorial on [this chapter](https://r4va.netlify.app/chap03). These are the R packages that are needed for plotting interactive graphs.
```{r}
pacman::p_load(ggiraph, plotly, 
               patchwork, DT, tidyverse)
```

```{r}
#| code-fold: true
#| code-summary: "Importing data"
#| warning: false

exam_data <- read_csv("../data/Exam_data.csv")
```

## 3.2 Interactivity with `ggiraph`

### 3.2.1 Tooltip effect with *tooltip* aesthetic

::: panel-tabset

## Using tooltip

The code chunk below will generate a dotplot that is interactive such that by hovering the mouse pointer on an data point of interest, the student’s ID will be displayed.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

# Step 1: Create an interactive version of ggplot2 geom.
p <- ggplot(data = exam_data,
            aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),  # Here's the tooltip
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)

# Step 2: Generate an svg object to be displayed on an html page.
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618
)
```

## Customizing tooltip content

Hover over the data point and the tooltip will display the ID and Class of the student, instead of just the student ID in the previous tab. 

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

# Step 1: Creating a new column in the dataset to incorporate the tool tip information
exam_data$tooltip <- c(paste0(
  "Name = ", exam_data$ID,
  "\n Class = ", exam_data$CLASS
))

# Step 2: Create an interactive version of ggplot2 geom.
p <- ggplot(data = exam_data,
            aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = exam_data$tooltip),  # Here's the tooltip
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)

# Step 3: Generate an svg object to be displayed on an html page.
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618
)
```

## Customizing tooltip style

This changes the tooltip from a black background + white text to a white background + black, bold text.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

# Step 1: Define tooltip CSS style
tooltip_css <- "background-color:white;
font-style:bold; color:black;"

# Step 2: Create an interactive version of ggplot2 geom.
p <- ggplot(data = exam_data,
            aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),  # Here's the tooltip
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)

# Step 3: Generate an svg object to be displayed on an html page.
girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6*0.618,
  options = list(  
    opts_tooltip(
      css = tooltip_css  # The tooltip css is incorporated here
    )
  )
)
```

:::

The code chunk below is an example of customizing the tooltip to show the 90% confidence interval.
```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

tooltip <- function(y, ymax, accuracy = .01) {
  mean <- scales::number(y, accuracy = accuracy)
  sem <- scales::number(ymax - y, accuracy = accuracy)
  paste("Mean maths scores:", mean, "+/-", sem)
}

gg_point <- ggplot(data = exam_data, 
                   aes(x = RACE),
) +
  stat_summary(aes(y = MATHS, 
                   tooltip = after_stat(  
                     tooltip(y, ymax))),  
    fun.data = "mean_se", 
    geom = GeomInteractiveCol,  
    fill = "light blue"
  ) +
  stat_summary(aes(y = MATHS),
    fun.data = mean_se,
    geom = "errorbar", width = 0.2, size = 0.2
  )

girafe(ggobj = gg_point,
       width_svg = 8,
       height_svg = 8*0.618)
```

### 3.2.2 Hover effect with *data_id* aesthetic

::: panel-tabset

## Hover effect

Elements associated with the data_id (i.e CLASS) will be highlighted upon mouse over.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

p <- ggplot(data = exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(           
    aes(data_id = CLASS),   # Here is the data_id          
    stackgroups = TRUE,               
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618                      
)         
```

## Styling hover effect

The css codes are used to change the highlighting effect. 
```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

p <- ggplot(data = exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(data_id = CLASS),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(                        
    opts_hover(css = "fill: #202020;"),  # The elements associated with the data_id
    opts_hover_inv(css = "opacity:0.2;")  # The ones that are not highlighted
  )                                        
)            
```

:::

The code chunk below combines what we have learnt so far - the tooltip + hover effect.
```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

p <- ggplot(data = exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(tooltip = CLASS, 
        data_id = CLASS),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618,
  options = list(                        
    opts_hover(css = "fill: #202020;"),  
    opts_hover_inv(css = "opacity:0.2;") 
  )                                        
)            
```

### 3.2.3 Onclick effect with `onclick`

The `onclick` argument of ggiraph provides hotlink interactivity on the web. The web document link with a data object will be displayed on the web browser upon mouse click.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

exam_data$onclick <- sprintf("window.open(\"%s%s\")",
"https://www.moe.gov.sg/schoolfinder?journey=Primary%20school",
as.character(exam_data$ID))

p <- ggplot(data = exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(onclick = onclick),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +               
  scale_y_continuous(NULL,               
                     breaks = NULL)
girafe(                                  
  ggobj = p,                             
  width_svg = 6,                         
  height_svg = 6*0.618)   
```

::: {.callout-warning}
Note that click actions must be a string column in the dataset containing valid javascript instructions.
:::

### 3.2.4 Coordinated multiple views
Coordinated multiple views methods has been implemented in the data visualisation below. When a data point of one of the dotplot is selected, the corresponding data point ID on the second data visualisation will be highlighted too. The steps will be annotated in the code chunk.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

# Step 1: Plot the first figure
p1 <- ggplot(data = exam_data, 
       aes(x = MATHS)) +
  geom_dotplot_interactive(              
    aes(data_id = ID,
        tooltip = ID),              
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") +  
  coord_cartesian(xlim = c(0,100)) + 
  scale_y_continuous(NULL,               
                     breaks = NULL)

# Step 2: Plot the second figure
p2 <- ggplot(data = exam_data, 
       aes(x = ENGLISH)) +
  geom_dotplot_interactive(              
    aes(data_id = ID,
        tooltip = ID),        
    stackgroups = TRUE,                  
    binwidth = 1,                        
    method = "histodot") + 
  coord_cartesian(xlim = c(0,100)) + 
  scale_y_continuous(NULL,               
                     breaks = NULL)

# Step 3: Display the svg object. The code argument runs the patchwork function to create the coordinated multiple views.
girafe(code = print(p1 + p2),  # print is needed for the plot to display
       width_svg = 6,
       height_svg = 3,
       options = list(
         opts_hover(css = "fill: #202020;"),
         opts_hover_inv(css = "opacity:0.2;")
         )
       ) 
```

### 3.2.5 Extra: Modification from previous exercise
Last week, we learnt to use the `ggrepel` to prevent overlapping text labels [here](week02.html#beyond-ggplot2-annotation-ggrepel). However, the plot was still overwhelming. 

::: panel-tabset

## With ggrepel
```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

ggplot(data = exam_data, 
       aes(x = MATHS, 
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm, 
              linewidth = 0.5) +  
  coord_cartesian(xlim = c(0,100),
                  ylim = c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## Using hover instead
```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

p <- ggplot(data = exam_data, 
       aes(x = MATHS, 
           y = ENGLISH)) +
  geom_point_interactive(
    aes(tooltip = ID)
  ) +
  geom_smooth(method = lm, 
              linewidth = 0.5) +  
  coord_cartesian(xlim = c(0,100),
                  ylim = c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")

girafe(                                  
  ggobj = p) 
```

:::

## 3.3 Interactivity with `plotly`

### 3.3.1 Creating an interactive scatterplot with `plot_ly()`

::: panel-tabset

## Default

With just three lines of code, we can plot a basic interactive plot.
```{r}
#| warning: false

plot_ly(data = exam_data, 
        x = ~MATHS, 
        y = ~ENGLISH)
```

## Using color
```{r}
#| warning: false

plot_ly(data = exam_data, 
        x = ~MATHS, 
        y = ~ENGLISH,
        color = ~RACE)
```

:::

### 3.3.2 Creating an interactive scatterplot with `ggplotly()`

To create an interactive scatterplot, all we need to do is to wrap the figure using `ggplotly()`.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

p <- ggplot(data = exam_data,
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size = 1) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100))

ggplotly(p)
```

### 3.3.3 Coordinated multiple views

Expand on the code chunk to see the steps needed to create this plot. 
```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

# Step 1: Using highlight_key function from plotly package to indicate the shared data
# This creates an object of class crosstalk::SharedData
d <- highlight_key(exam_data)

# Step 2: Create both the scatterplots
p1 <- ggplot(data = d, 
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

p2 <- ggplot(data = d, 
            aes(x = MATHS,
                y = SCIENCE)) +
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

# Step 3: Using subplot to place them side-by-side
subplot(ggplotly(p1),
        ggplotly(p2))
```

On combining multiple views, refer to this [link](https://plotly-r.com/arranging-views) to learn more.

## 3.4 Interactivity with `crosstalk`

The `DT::datatable` can be used to render the data objects in R as HTML tables. The interactive table can then be combined with the ggplot graphics using `crosstalk`. The code chunk below is used to implement the coordinated brushing shown.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

# Step 1: Indicate shared data
d <- highlight_key(exam_data) 

# Step 2: Define the ggplot graphic
p <- ggplot(d, 
            aes(ENGLISH, 
                MATHS)) + 
  geom_point(size=1) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

# Step 3: Using the plotly highlight function
gg <- highlight(ggplotly(p),        
                "plotly_selected")  

# Step 4: Putting HTML elements side by side.
crosstalk::bscols(gg,               
                  DT::datatable(d), 
                  widths = 5)  
```

# 4 Programming Animated Statistical Graphics with R

## 4.1 Introduction

The basic concepts of animation and the terminologies are covered in the chapter [here](https://r4va.netlify.app/chap04). For me, the two key takeaways are:

1. **What We Can Do with Animated Graphics** <br>
Animated graphics help create engaging and impactful data visualizations by showing changes over time or across categories. They are built by stitching together multiple frames, much like a flipbook, using tools like `gganimate` and `plotly` in R. Additionally, data reshaping (`tidyr`) and transformation (`dplyr`) are essential for preparing data for animation. Animation attributes, such as frame duration and easing functions, allow further customization of how the animation flows.

2. **When to Use Animated Graphics** <br>
Animated graphics are most effective in presentations or storytelling when visualizing trends or patterns over time. Their ability to guide audience attention makes them more engaging than static visuals. However, they may not be necessary for exploratory data analysis, where static charts might suffice. Always consider whether animation adds value before investing time in creating it.

```{r}
#| warning: false
pacman::p_load(readxl, gifski, gapminder,
               plotly, gganimate, tidyverse)
```

```{r}
#| code-fold: true
#| code-summary: "Importing data"
#| warning: false

# Step 1: Select the columns to be converted into factor (levels)
col <- c("Country", "Continent")

# Step 2: Read xls from readxl to import excel worksheet
# Step 3: Piping the operators together, using mutate_at & mutate for datatype conversion
globalPop <- read_xls("../data/GlobalPopulation.xls",
                      sheet="Data") %>%
  mutate_at(col, as.factor) %>%  # mutate(across(col, as.factor)) can also be used
  mutate(Year = as.integer(Year))
```

## 4.2 Animated Data Visualisation: `gganimate` methods

::: panel-tabset
## Static
```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

ggplot(data = globalPop, 
       aes(x = Old, 
           y = Young, 
           size = Population, 
           colour = Country)) +
  geom_point(alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}', 
       x = '% Aged', 
       y = '% Young') 
```

## Animated
```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

ggplot(data = globalPop, 
       aes(x = Old, 
           y = Young, 
           size = Population, 
           colour = Country)) +
  geom_point(alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}', 
       x = '% Aged', 
       y = '% Young') +
  transition_time(Year) +  # Transition through distinct states in time   
  ease_aes('linear')  # Control of easing of aesthetics
```
:::

## 4.3 Animated Data Visualisation: plotly

### 4.3.1 Animated bubble plot with `ggplotly()` method

::: panel-tabset

## Default

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

gg <- ggplot(data = globalPop, 
       aes(x = Old, 
           y = Young, 
           size = Population, 
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7, 
             show.legend = FALSE) +  # The legend is still displayed
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged', 
       y = '% Young')

ggplotly(gg)
```

The animated bubble plot above includes a play/pause button and a slider component for controlling the animation. 

## Removing the legend

Although `show.legend = FALSE` argument was used in the default plot, the legend still appears.. To overcome this problem, `theme(legend.position='none')` should be used as shown in the plot and code chunk below.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

gg <- ggplot(data = globalPop, 
       aes(x = Old, 
           y = Young, 
           size = Population, 
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged', 
       y = '% Young') +
  theme(legend.position = 'none')

ggplotly(gg)
```

:::

### 4.3.2 Animated bubble plot with `plot_ly()` method

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

bp <- globalPop %>%
  plot_ly(x = ~Old, 
          y = ~Young, 
          size = ~Population, 
          color = ~Continent,
          sizes = c(2, 100),
          frame = ~Year, 
          text = ~Country, 
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers') %>%
  layout(showlegend = FALSE)
bp
```

