---
title: "Week 01 - A Layered Grammar of Graphics: `ggplot2` methods"
format: html
sidebar: false
date-modified: last-modified
---

## 1.1 Introduction

The following content is created by following the tutorial on [this chapter](https://r4va.netlify.app/chap01).

```{r}
pacman::p_load(tidyverse)

exam_data <- read_csv("../data/Exam_data.csv")
```

### 1.1.1 Comparison of R Graphics VS ggplot

::: panel-tabset
## R Graphics

```{r}
hist(exam_data$MATHS, xlab="Math Score", main="Distribution of Math Scores")
```

## ggplot2

```{r}
# ggplot(data=df) will just display an empty canvas.
# setting aes(x=MATHS) just sets the x-axis to the range of values in the MATHS column
# using geom_histogram 
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10,
                 boundary=100,
                 color="black",
                 fill="grey") +
  ggtitle("Distribution of Math Scores")
```
:::

## 1.2 Geometric Objects in ggplot2: geom

### 1.2.1 Bar chart with `geom_bar()`

```{r}
ggplot(data=exam_data,
       aes(x=RACE)) +
  geom_bar()
```

### 1.2.2 Dot plot with `geom_dotplot()`

Not be confused with scatterplots, a dot plot is essentially a histogram but in the style of stacked dots.

```{r}
ggplot(data=exam_data, 
       aes(x=MATHS)) +
  geom_dotplot(dotsize=0.5)
```

Because the y-axis scale is misleading, we will have to turn off y-axis with the `scale_y_continuous()`

```{r}
ggplot(data=exam_data,
       aes(x=MATHS)) +
  geom_dotplot(binwidth=2.5,
               dotsize=0.5) +
  scale_y_continuous(NULL, breaks=NULL)
```

Question: When is a dot plot preferred over something like a histogram? Is counting the dots even feasible?

### 1.2.3 Histograms with `geom_histogram()`

```{r}
# default number of bins is 30
ggplot(data=exam_data,
       aes(x=MATHS)) +
  geom_histogram()
```

In the `geom()`, we can use the arguments to modify the geometric object.

```{r}
ggplot(data=exam_data,
       aes(x=MATHS)) +
  geom_histogram(bins=20,
                 color="black",
                 fill="light blue")
```

Similar to `hue` in `seaborn`, we can colour the histogram by sub-groups using the `fill` argument in `aesthetic()`. We can also play around with the other arguments like `alpha` and `color`.

::: panel-tabset
## `fill`

```{r}
ggplot(data=exam_data,
       aes(x=MATHS,
           fill=GENDER)) +
  geom_histogram(bins=20,
                 color="grey30")
```

## `alpha`

```{r}
# Can't use it on continuous variables either even though there's the warning.
# It's very difficult to see.
ggplot(data=exam_data,
       aes(x=MATHS,
           alpha=CLASS)) +
  geom_histogram(bins=20,
                 color="grey30")
```

## `color`

The `color` argument in the `geom_histogram()` needs to be removed as it will override the previous `aes()`

```{r}
ggplot(data=exam_data,
       aes(x=MATHS,
           color=GENDER)) +
  geom_histogram(bins=20,
                 alpha=0.3)
```
:::

### 1.2.4 Kernel Density Estimate with `geom_density()`

```{r}
ggplot(data=exam_data,
       aes(x=MATHS)) +
  geom_density()

ggplot(data=exam_data,
       aes(x=MATHS,
           color=GENDER)) +
  geom_density()
```

### 1.2.5 Boxplot with `geom_boxplot()`

```{r}
ggplot(data=exam_data,
       aes(y=MATHS,
           x=GENDER)) +
  geom_boxplot()

# notches are used to see if medians of the distributions differ
ggplot(data=exam_data,
       aes(y=MATHS,
           x=GENDER)) +
  geom_boxplot(notches=TRUE) # TRUE must be all caps
```

### 1.2.6 Violin plot with `geom_violin()`

This is for comparing multiple data distributions side by side. With density curves, as the lines may overlap with one another, it is more challenging to compare the different distributions.

```{r}
ggplot(data=exam_data, 
       aes(y=MATHS, 
           x= GENDER)) +
  geom_violin()
```

### 1.2.7 Scatterplot with `geom_point()`

```{r}
ggplot(data=exam_data,
       aes(x=MATHS,
           y=ENGLISH)) +
  geom_point()
```

### 1.2.8 Combining *geom* objects

```{r}
ggplot(data=exam_data,
       aes(y=MATHS,
           x=GENDER)) +
  geom_boxplot() +
  geom_point(position="jitter",
             size=0.5)
```

## 1.3 Statistics functions in `ggplot2`: stat

### 1.3.1 Using the `stat_summary()` method

The box plot above doesn't indicate the position of the means. By working with the `stat_summary()` function, we can indicate the mean values on the plot.

```{r}
ggplot(data=exam_data,
       aes(y=MATHS,
           x=GENDER)) +
  geom_boxplot() +
  stat_summary(geom="point",
                fun="mean",
                colour="red",
                size=4)
```

### 1.3.2 Using `geom()` to update stat

Overriding the default stat, we can add the mean values as well.

```{r}
ggplot(data=exam_data,
       aes(y=MATHS,
           x=GENDER)) +
  geom_boxplot() +
  geom_point(stat="summary",
             fun="mean",
             colour="red",
             size=4)
```

### 1.3.3 Using `geom_smooth()` to add best fit curve

::: panel-tabset
## loess

```{r}
ggplot(data=exam_data, 
       aes(x=MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(linewidth=0.5)
```

## lm

```{r}
ggplot(data=exam_data, 
       aes(x=MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm,
              linewidth=0.5)
```
:::

## 1.4 Facets

Also known as trellis plot. In mpl it's subplots.

### 1.4.1 Wrapping 1d sequence of panels into 2d with `facet_wrap()`

```{r}
ggplot(data=exam_data,
       aes(x=MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

### 1.4.2 Creating matrix of panels with `facet_grid()`

```{r}
# This should be similar to subplots
ggplot(data=exam_data,
       aes(x=MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

## 1.5 Coordinates

### 1.5.1 Flipping the coordinates with `coord_flip()`

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

### 1.5.2 Changing the y- and x-axis range with `coord_cartesian()`

```{r}
ggplot(data=exam_data, 
       aes(x=MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

## 1.6 Themes

-   [List](https://ggplot2.tidyverse.org/reference/ggtheme.html) of built-in themes

### 1.6.1 Examples of different themes

::: panel-tabset
## theme_gray()

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

## theme_classic()

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

## theme_minimal()

```{r}
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```
:::

## 1.7 Additional Plots
This space is for more exploration regarding this week's extra topics. It may be done later when I have time to revise. 

### 1.7.1 Age-sex Pyramid
Data Retrieved From [Singstat](https://tablebuilder.singstat.gov.sg/table/TS/M810011#!)

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

# Referenced from https://www.geeksforgeeks.org/how-to-create-a-population-pyramid-in-r/
pop_data <- read_csv("../data/sg_residents_jun2024.csv")

# This is to ensure that the age levels are sorted correctly
age_levels <- pop_data %>%
  distinct(age) %>%
  pull(age) %>%
  str_replace(" Years", "")
  
pop_data <- pop_data %>%
  mutate(age = factor(str_replace(age, " Years", ""), levels = age_levels))

# Plotting the age-sex pyramid
pop_data %>%mutate( 
    population = ifelse(gender=="M", population*(-1e-3), 
                        population*1e-3))%>% 
    ggplot(aes(x = age, 
               y = population, 
               fill = gender)) +  
    geom_bar(stat = "identity") + 
    coord_flip() + 
    scale_y_continuous(limits = c(-200,200),  
                       breaks = seq(-200, 200, by = 50))+
   labs(title = "Singapore Residents", x = "Age Range",  
        y = "Population (in thousands)") +
  theme_light() +
  theme(axis.ticks.y=element_blank())
```

### 1.7.2 Pareto Chart
Data Retrieved From [Singstat](https://tablebuilder.singstat.gov.sg/table/TS/M212981)

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

pacman::p_load(scales)

exp_data <- read_csv("../data/avg_monthly_expenditure2023.csv")

pareto_data <- exp_data %>%
  mutate (
    category_wrapped = str_wrap(category, width = 8)
  ) %>% 
  arrange(desc(avg_spend)) %>%  # Sort by avg_spend in descending order
  mutate(
    category_wrapped = factor(category_wrapped, levels = category_wrapped),  # Fix category order
    cum_percent = cumsum(avg_spend) / sum(avg_spend) * 100)  # Calculate cumulative percentage


# Create the Pareto chart
ggplot(pareto_data, aes(x = category_wrapped)) +
  geom_bar(aes(y = avg_spend), stat = "identity", fill = "steelblue") +
  geom_line(aes(y = cum_percent * max(avg_spend) / 100, 
                group = 1), color = "red", 
            size = 1) +
  geom_point(aes(y = cum_percent * max(avg_spend) / 100), 
             color = "red", size = 2) +
  scale_y_continuous(
    name = "Average Monthly HouseHold Expenditure (SGD)", 
    sec.axis = sec_axis(~ . * 100 / max(pareto_data$avg_spend),
                        breaks = seq(0, 100, by = 20),
                        labels = percent_format(scale = 1))) +
  labs(
    title = "Average Monthly Household Expenditure by Goods and Services Type (2023)",
    x = "Type of Goods and Services",
    y = "Average Monthly HouseHold Expenditure (SGD)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6)) 
```


This Pareto chart shows the average monthly household expenditure in Singapore for 2023, offering insights into spending priorities. Key areas like "Housing and Utilities," "Food and Food Serving Services," and "Transport" form the bulk of household expenses. However, it’s worth noting that the data is self-reported, as outlined in the Household Expenditure Survey 2022/23 ([SingStat](https://www.singstat.gov.sg/hes/hes-2223)). Additionally, this analysis isn’t normalized by household size or segmented by income levels, which could mask inequalities across different demographics. 

For a deeper dive, the full report and related publications are available on SingStat’s website ([here](https://www.singstat.gov.sg/publications/households/household-expenditure-survey)). This serves as a starting point for exploring how households allocate resources, but there’s room to uncover more patterns with richer, disaggregated data.